A formal approach but practical approach in train regulation

### 2025-06-23 paper/main.tex
composition 1

%% Modèle
\newcommand{\modelSet}{\ensuremath{\mathbb{M}}}

@@

%% Switches
\definecolor{swcolor}{rgb}{0.4, 0.23, 0.56}
\newcommand{\swFmt}[1]{{\color{swcolor}{#1}}}
\newcommand{\sid}[1]{\ensuremath{\swFmt{s_{#1}}}}
\newcommand{\turnouts}{\swFmt{\mathsf{\Theta}}}
\newcommand{\switches}{\ensuremath{\swFmt{\sigma}}}
\newcommand{\networkConf}{\ensuremath{\swFmt{\Sigma}}}
\newcommand{\nosuc}{\ensuremath{\swFmt{\bot}}}
\newcommand{\updateSignals}[3]{\ensuremath{#1[\swFmt{#2}\leftarrow #3]}}

@@

%% Regulator
\newcommand{\handlerOf}[2]{\ensuremath{#1[\trainFmt{#2}]}}
\newcommand{\popHandlerHead}[2]{\ensuremath{{\tt pop}(#1[\trainFmt{#2}])}}
\newcommand{\substHandlerHead}[3]{\ensuremath{#1[\trainFmt{#2} \leftarrow #3]}}

%% Regulator's Program
\newcommand{\regTuple}[3]{\tuple{#1, #2, #3}}
\newcommand{\evGuard}[2]{\ensuremath{\mathtt{ev}(\trainFmt{#1}, #2)}}
\newcommand{\ev}[3]{\evGuard{#1}{#2}: #3}
\newcommand{\incr}[1]{{\mbox{\ensuremath{\mathtt{incr}(\posFmt{#1})}}}\xspace}
\newcommand{\turn}[2]{{\mbox{\ensuremath{\mathtt{turn}(#1, #2)}}}\xspace}
\newcommand{\tokens}{\ensuremath{T}}
\newcommand{\tokenOf}[1]{\ensuremath{T(\posFmt{#1})}}
\newcommand{\regulator}{\ensuremath{R}}
\newcommand{\wait}[2]{{\mbox{\ensuremath{\mathtt{wait}(\posFmt{#1}, #2)}}}\xspace}
\newcommand{\regConcat}[2]{#1 \cdot #2}

@@

%% Buffer
\newcommand{\bufferFmt}[1]{#1}
\newcommand{\head}[1]{\ensuremath{{\tt hd}(#1)}}
\newcommand{\buflen}[1]{\ensuremath{{\tt len}(\bufferFmt{#1})}}
\newcommand{\bufhead}[1]{\head{\bufferFmt{#1}}}
\newcommand{\buftail}[1]{\ensuremath{{\tt tl}(\bufferFmt{#1})}}
\newcommand{\emptyList}{\ensuremath{\varepsilon}}
\newcommand{\bufferSet}{\ensuremath{\bufferFmt{\mathbb{B}}}\xspace}

@@

\begin{itemize}
	\item The routing program $E$ specifies how the regulator should react to events (i.e. when a train enters a block). A program is a collection of rules to apply upon each event (called an \emph{event handler}). We detail further event handlers below.
	\item The token list, denoted $\tokens$, is a function  $\tokens: \blocks \to \Nat$ associating a token level to each block. Those tokens are used to prevent multiple trains accessing the same block at the same time.
	\item The waiting set is a set $W \subseteq \blocks \times \Nat \times \trains$. A triple $\tuple{\bid{id}, n, \tid{id}} \in W$ means block $\bid{id}$ must be protected against train $\tid{id}$ entering as long as $\tokenOf{\bid{id}} < n$.
\end{itemize}

@@

\paragraph{Regulator's event handlers.}

Train events are handled by the regulator. Upon such events, the regulator can take actions such as turning a switch, allowing or blocking a train, etc.. The regulator is equipped with a routing program. This routing program contains, for each train, a list of handlers. For a given train, the first handler of the list is the set of actions that have to be executed on the train's first event, second handler for the second event, and so on.

Given a routing program $E$, $\handlerOf{E}{\tid{id}}$ is the list of handlers of train $\tid{id}$, $\head{\handlerOf{E}{\tid{id}}}$ is the handler of the first event for that train.

The syntax of each handler is given by the non-terminal {\ttfamily stmt} in Grammar~\ref{gr:regulator_program}.\todo{est-ce toujours nécessaire maintenant que les handlers sont dans un liste ?}
The program of the regulator is a set of rules, guarded by events. A rule $\ev{\tid{id}}{n}{A}$ denotes that upon entering the $n$-th block of its route, train $\tid{id}$ will trigger actions $A$. The list of actions in the handler are of four kinds: 
\begin{description}
	\item[\incr{\bid{id}}:] this updates the token $\bid{id}$ in $T$:  $T \cdot \incr{\bid{id}}  \triangleq T[\bid{id} \leftarrow T(\bid{id})+1]$
	\item [\wait{\bid{id}}{i}:] ...\todo{}
	\item [\turn{\sid{id}}{s}:] this instructs the regulator to turn the switch $\sid{id}$ into the $s$ state: $\sigma\cdot\turn{\sid{id}}{s} = \sigma[\sid{id} \leftarrow s]$
	\item [auth:] ...\todo{}
\end{description}

Notice that all actions of a rule are executed atomically. Our semantics, presented below, details actions' behaviour.
\todo{Clarifier Event, indexation par train, qui donne une liste de handlers, eux-même une liste d'ordres}

\begin{grammar}[Regulator's rules, with $\tid{id} \in \trains$, $n \in  \Nat$, $\bid{id} \in \blocks$, $\sid{id} \in \turnouts$, $s\in \internalState$.][h][gr:regulator_program]
	\firstcasesubtil{\ttfamily stmt}{\evGuard{\tid{id}}{n} : \nonterm{A}^\star}{}
	\firstcase{A}{\incr{\bid{id}}}{}
	\otherform{\wait{\bid{id}}{i}}{}
	\otherform{\turn{\sid{id}}{s}}{}
	\otherform{auth}{}
\end{grammar}

\paragraph{Model.}
Overall, states of our model are tuples $\tuple{\trainSeq, \regulator, \switches, \signals}$ where $\trainSeq$ is a set of train tuples, $\regulator$ a regulator tuple, $\switches$ a network configuration (indicating trackworks states) and $\signals$ a signals state function. We suppose all train tuples in $\trainSeq$ have a different identifier. Notice that $\sucblock$ is not part of the state. Instead, the dynamics of the model (see below) is parameterised by $\sucblock$. Let $\modelSet$ be the set of all model states.

@@

In the previous section, we introduced our model. In this section, we present the dynamics of our model, that is the rules that govern how our model can switch from one state to the other.
We enrich our model with two elements: a guard, which goal is to schedule agents (for instance, it ensures the regulator acts atomically); and a communication buffer, which stores pending events until the regulator treats them. We present those two new elements in the first subsection.
The dynamics is described by means of a reduction relation, characterising the successive states of our model. We define this reduction relation using inference rules in the second subsection.

@@

Train events are not dealt with instantly by the regulator (e.g. two train events can happen before the regulator begins to react to the first one). To characterise this behaviour, train events are stored in a buffer until the regulator uses them. Train events are tuples \tuple{\tid{id}, \bid{id}}, indicating that the train with id $\tid{id}$ enters block \bid{id}. A buffer \bufferFmt{B} is a sequence of such events. Let \bufferSet{} be the set of all buffers.

@@

States of the system are triples from $\{\guardR, \guardS, \guardT\} \times \bufferSet \times \modelSet$. Our reduction relation is a binary relation on system states, noted $\reduces$: $S_1 \reduces S_2$ means that state $S_1$ can change to $S_2$ (note that this relation is not a function: $S_1$ can reduce to multiple states, due to non-determinism). We define $\reduces$ as the least relation induced by the inference rules presented hereafter.

@@

\begin{mathpar}
	\inferrule*[left=\ruleDef{Start}]{
		\dirFmt{d} \neq \dirFmt{d^\prime} \and P = \trainConcat{\su{d^\prime}{\posFmt{N}}}{P^\prime}
	}{
		\hspace{3mm} \redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{P}\}}{\regulator}{\switches}{\signals}} \\
		\reduces
		\redTuple{\guardT}{\push{B}{\tuple{\tid{id}, \bid{id}}}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d^\prime}{P}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}

Conversely, if the train program is empty (i.e. upon termination), the train stops using rule \ruleRef{Stop}.
\begin{mathpar}
	\inferrule*[left=\ruleDef{Stop}]{
		\dirFmt{d} \neq \dirStop
	}{
	\hspace{4mm}  \redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{\emptyTrainProg}\}}{\regulator}{\switches}{\signals}} \\
	\reduces
	\redTuple{\guardT}{\push{B}{\tuple{\tid{id}, \bid{id}}}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{\dirStop}{\emptyTrainProg}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}

@@

\begin{mathpar}
	\inferrule*[left=\ruleDef{UntilNext}] {
		\suc{\bid{id}}{d}{\switches} = \bid{id^\prime} \neq \nosuc
		\and
		\signalF{\bid{id}}{d} = \siggreen
	}{
		\redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{\trainConcat{\su{d}{[\bid{id^\prime}]}}{P}}\}}{\regulator}{\switches}{\signals}} \\
		\reduces		
		\redTuple{\guardS}{\push{\bufferFmt{B}}{\tuple{\tid{id}, \bid{id^\prime}}}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id^\prime}}{d}{P}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}\todo{Pas de message dans ce cas là?}

@@

When the regulator reacts to a train event, it executes atomically all rules associated with that event. To implement this atomic execution, the first regulator event is to change the guard to \guardR (see Fig. \ref{fig:state_machine_guard}).

\begin{mathpar}
	\inferrule*[left=\ruleDef{StartEvent}] {
		\buflen{B} \neq 0
	}{
		\tuple{\guardT, B, R} \reduces \tuple{\guardR, B, R}
	}
\end{mathpar}

Then, the regulator lookup the first event on the buffer, and executes all the rules in the event handler associated with this event. All actions are treated in order, and when the list of remaining actions is empty, the handling terminates and the event is removed of the buffer. We introduce 6 reduction rules for the 4 possible kinds of orders (\incr{\bid{id}} and \wait{\bid{id}}{i} have two rules).

Rule \ruleRef{Turn} triggers when a \turn{\sid{id}}{s} is encountered in the event handler. It updates the state of the targeted switch.
\begin{mathpar}
	\inferrule*[left=\ruleDef{Turn}]{
		\bufhead{B} = \tuple{\tid{id}, \bid{id}}
		\and
		\head{\handlerOf{E}{\tid{id}}} = \regConcat{\turn{\sid{id}}{s}}{A^\prime}
	}{
	\redTuple{\guardR}{\bufferFmt{B}}{\stateTuple{\trainSeq}{\tuple{E, T, W}}{\switches}{\signals}}
	\reduces
	\redTuple{\guardR}{\bufferFmt{B}}{\stateTuple{\trainSeq}{\tuple{\substHandlerHead{E}{\tid{id}}{A^\prime}, T, W}}{\updateSignals{\switches}{\sid{id}}{s}}{\signals}}
	}
\end{mathpar}

\todo{other rules}

Finally, when the handler of the pending event has no action leftover, the regulator terminates its handling: it removes the (now empty) head of $\handlerOf{E}{\tid{id}}$, the head of the buffer, and returns the guard to state $\guardT$.

\begin{mathpar}
	\inferrule*[left=\ruleDef{EndEvent}]{
		\bufhead{B} = \tuple{\tid{id}, \bid{id}}
		\and
		\head{\handlerOf{E}{\tid{id}}} = \emptyList
	}{
		\redTuple{\guardR}{\bufferFmt{B}}{\stateTuple{\trainSeq}{\tuple{E, T, W}}{\switches}{\signals}}
		\reduces
		\redTuple{\guardT}{\buftail{B}}{\stateTuple{\trainSeq}{\tuple{\popHandlerHead{E}{\tid{id}}, T, W}}{\switches}{\signals}}
	}
\end{mathpar}

@@

Variables will update with time, each transition is described with prime variables. Those represent the value of the variable after the transition. For instance, \mintinline{C}{gamma' = [gamma EXCEPT ![T.id].d = "f"]} updates the train tuple: the train \mintinline{C}{T.id}'s direction is set to \mintinline{C}{"f"} where  \mintinline{C}{T} is a variable  defined somewhere else.

To each rule in Section~\ref{sec:formal-model}, we associate a TLA+ rule taking the shape (here the Stop rule): 
\begin{minted}{C}
Stop (T) ==
        /\ Len(T.prog) = 0
        /\ meta.G.state = "none"
        /\ T.d /= "*"
        /\ gamma' = [gamma EXCEPT ![T.id].d = "*"]
        /\ meta' = [meta EXCEPT !.msg[1] = Append(meta.msg[1],<<T.id,T.pid>>)]
        /\ UNCHANGED << reg, sigma, F >>
 \end{minted}
The first three lines correspond to the conditions that will fire the rule, the last ones correspond to the configuration update. The next definition represent the application of one rule:
\begin{minted}{C}
Next == 
   \/ UpdateTL
   \/ StartEvent
   \/ Turn
   ...
   \/  \E i \in 1..Len(gamma) :
            \/ Start(gamma[i])
            \/ Until(gamma[i])
            ...
\end{minted}

The system is then reprensented by a formula:
\begin{verbatim}
    Spec == Init /\  [][Next]_variables
\end{verbatim}%Revoir la syntaxe au cas où / remplacer par le jolie tla2latex

Actually, in our formal semantics, nothing forces a rule to be applied. As a consequence, an ongoing train may never reach its next block. Thanks to TLA+, we can ensure that a rule will eventually be applied (if it can be). Those are known as weak fairness conditions. Again, we specify it as follows. \todo{WF condition?}

@@

La composition est un procédé important qui nous permet de créer des chemins complexes à partir de chemins élémentaires.
Pour cela, nous décomposons cette fonction en trois parties : préparation, assemblage et nettoyage.
\\Notation : on note $o_x$ une liste d'ordres, avec $x \in \Nat$.

\subsection{Préparation}
\textbf{Simplification des ordres "turn"}\\
La préparation est une étape cruciale, nous permettant de simplifier la suite du processus.
D'abord, nous allons simplifier les ordres "turn", pour cela, {\color{red} on pose l'hypothèse que o1 et o2 sont élémentaires (embêtant si on veut faire des compositions complexes)}.
Ainsi, dans chaque liste d'ordre, un ordre "turn" ne peut concerner qu'un seul switch. On cherche alors les ordres similaires entre les listes pour les supprimer de o2.
\\Par exemple :

\begin{center}
	\begin{tabular}{||c || c | c||}
		\hline
		tid & o$_1$ & o$_2$ \\ [0.5ex] 
		\hline\hline
		0 & \tuple{10d,{\color{blue}11v}} & \tuple{} \\ 
		1 & \tuple{} & \tuple{{\color{blue}12v},10d} \\
		\hline
		0 & \tuple{{\color{blue}10d},11v} & \tuple{} \\ 
		1 & \tuple{} & \tuple{{\color{blue}12v},10d} \\
		\hline
		0 & \tuple{10d,{\color{violet}11v}} & \tuple{} \\ 
		1 & \tuple{} & \tuple{12v,{\color{violet}10d}} \\
		\hline
		0 & \tuple{\underline{\color{violet}10d},11v} & \tuple{} \\ 
		1 & \tuple{} & \tuple{12v,\underline{\color{violet}10d}} \\
		\hline
		0 & \tuple{10d,11v} & \tuple{} \\ 
		1 & \tuple{} & \tuple{12v} \\
		\hline
	\end{tabular}
\end{center}
On convertie ensuite les ordres "turn" restant dans o1 en "initialisation" d'un aiguillage si ce dernier n'est pas déjà initialisé.
\\\textbf{Traiter les ressources critiques}\\
Pour cette étape, on cherche d'abord à identifier les ressources critiques. Avec une méthode similaire à celle vue précédemment :

\vspace{0.5cm}

\begin{center}
	\begin{tabular}{||c || c | c||} 
		\hline
		tid & o$_1$ & o$_2$ \\ [0.5ex] 
		\hline\hline
		0 & \tuple{4,3,{\color{red}8}} & \tuple{8} \\ 
		1 & \tuple{6} & \tuple{{\color{red}7},3,4,5} \\
		\hline
		0 & \tuple{4,{\color{red}3},8} & \tuple{8} \\ 
		1 & \tuple{6} & \tuple{{\color{red}7},3,4,5} \\
		\hline
		0 & \tuple{{\color{red}4},3,8} & \tuple{8} \\ 
		1 & \tuple{6} & \tuple{{\color{red}7},3,4,5} \\
		\hline

		0 & \tuple{4,3,{\color{violet}8}} & \tuple{8} \\
		1 & \tuple{6} & \tuple{7,{\color{violet}3},4,5} \\
		\hline
		0 & \tuple{4,\underline{\color{violet}3},8} & \tuple{8} \\ 
		1 & \tuple{6} & \tuple{7,\underline{\color{violet}3},4,5} \\
		\hline

		0 & \tuple{4,3,{\color{blue}8}} & \tuple{8} \\
		1 & \tuple{6} & \tuple{7,3,{\color{blue}4},5} \\
		\hline

		0 & \tuple{4,3,{\color{orange}8}} & \tuple{8} \\
		1 & \tuple{6} & \tuple{7,3,4,{\color{orange}5}} \\
		\hline
	\end{tabular}
\end{center}
\noindent
Dans notre exemple, on peut voir que 3 est la seule ressource critique, il faut donc traiter ce cas.
On va ajouter un ordre "att(3,valT)" à l'event précédent l'entrée de la ressource critique pour le train qui utilise la ressource en second.
On ajoute également un ordre "incr(3)" à l'event suivant la sortie de la ressource critique pour le train qui utilise la ressource en premier.
Si le train à besoin d'un "turn" pour entrer dans la ressources critique, alors on le décale dans un event du train ayant utilisé l'aiguillage le plus récemment.
\\\textbf{Mettre à jour les jetons}\\
Laissons de côté l'hypothèse que o1 et o2 sont élémentaires, il est alors possible que o1 contienne des ordres de type "att" et "incr". Il faut donc mettre à jour les valeurs
des tokens pour "att" et "incr" dans o2. Par exemple, si dans o1 on trouve les ordres : (att(3,1), incr(3)), alors dans o2 on va devoir transformer att(3,1) en att(3,2).

\subsection{Assemblage}
L'assemblage est la partie la plus simple de cette fonction, il s'agit de concaténer les ordres des trains et des events.
Pas de technique intelligente, si les programmes du train 1 pour o1 et o2 sont "<L,[1,2]>" et "<L,[3,4]>", alors la concaténation donne "<L,[1,2]>,<L,[3,4]>".
Même chose pour les events, \tuple{\tuple{1,\tuple{}},\tuple{2,\tuple{turn(x,y)}}} et \tuple{\tuple{2,\tuple{att(3,1)}},\tuple{3,\tuple{}}} devient \tuple{\tuple{1,\tuple{}},\tuple{2,\tuple{turn(x,y),(att(3,1))}},\tuple{3,\tuple{}}}.
À partir de maintenant, nous ne manipulerons plus deux listes d'ordres, o1 et o2, mais une seule : o3. 
\\Aussi, on reprend l'initialisation des aiguillages de o1 pour o3.

\subsection{Nettoyage}
Cette partie fait office de "finalisation". On va d'abord s'occuper de "regrouper" les ordres des trains. Notre exemple précédent "<L,[1,2]>,<L,[3,4]>", nous donne maintenant "<L,[1,2,3,4]>". 
On ajoute également un ordre "auth" pour les "turn" utilisés par des trains pour agir sur leur propre parcours. (les trains n'utilisent pas directement "turn", mais plus simple à expliquer de cette manière)
Enfin, on chercher le premier arrêt de chaque train pour assigner les feux au rouge.
\section{Expériences}
\label{sec:experiments}

>>>>

@@

%% Switches
\definecolor{swcolor}{rgb}{0.4, 0.23, 0.56}
\newcommand{\swFmt}[1]{{\color{swcolor}{#1}}}
\newcommand{\sid}[1]{\ensuremath{\swFmt{s_{#1}}}}
\newcommand{\turnouts}{\swFmt{\mathsf{\Theta}}}
\newcommand{\switches}{\ensuremath{\swFmt{\sigma}}}
\newcommand{\networkConf}{\ensuremath{\swFmt{\Sigma}}}
\newcommand{\nosuc}{\ensuremath{\swFmt{\bot}}}

@@

%% Regulator's Program
\newcommand{\regTuple}[3]{\tuple{#1, #2, #3}}
\newcommand{\evGuard}[2]{\ensuremath{\mathtt{ev}(\trainFmt{#1}, #2)}}
\newcommand{\ev}[3]{\evGuard{#1}{#2}: #3}
\newcommand{\incr}[1]{{\mbox{\ensuremath{\mathtt{incr}(\posFmt{#1})}}}\xspace}
\newcommand{\turn}[2]{{\mbox{\ensuremath{\mathtt{turn}(#1, #2)}}}\xspace}
\newcommand{\tokens}{\ensuremath{T}}
\newcommand{\tokenOf}[1]{\ensuremath{T(\posFmt{#1})}}
\newcommand{\regulator}{\ensuremath{R}}

@@

%% Buffer
\newcommand{\bufferFmt}[1]{#1}

@@

\begin{itemize}
	\item The routing program $E$ specifies how the regulator should react to events (i.e. when a train enters a block). A program is a sequence of rules, which are described below.
	\item The token list, denoted $\tokens$, is a function  $\tokens: \blocks \to \Nat$ associating a token level to each block. Those tokens are used to prevent multiple trains accessing the same block at the same time.
	\item The waiting set is a set $W \subseteq \blocks \times \Nat \times \trains$. A triple $\tuple{\bid{id}, n, \tid{id}} \in W$ means block $\bid{id}$ must be protected against train $\tid{id}$ entering as long as $\tokenOf{\bid{id}} < n$.
\end{itemize}

@@

\paragraph{Regulator's rules.}

The syntax of regulator programs is given by the non-terminal {\ttfamily stmt} in Grammar~\ref{gr:regulator_program}.
The program of the regulator is a set of rules, guarded by events. A rule $\ev{\tid{id}}{n}{A}$ denotes that upon entering the $n$-th block of its route, train $\tid{id}$ will trigger actions $A$. There are four kinds of actions: 
\begin{description}
	\item[\incr{\bid{id}}:] this updates the token $\bid{id}$ in $T$:  $T \cdot \incr{\bid{id}}  \triangleq T[\bid{id} \leftarrow T(\bid{id})+1]$
	\item [att(pid, v):] ...\todo{}
	\item [\turn{\sid{id}}{s}:] this instructs the regulator to turn the switch $\sid{id}$ into the $s$ state: $\sigma\cdot\turn{\sid{id}}{s} = \sigma[\sid{id} \leftarrow s]$
	\item [auth:] ...\todo{}
\end{description}

Notice that all actions of a rule are executed atomically. Our semantics, presented below, details actions' behaviour.

\begin{grammar}[Regulator's rules, with $\tid{id} \in \trains$, $n \in  \Nat$, $\bid{id} \in \blocks$, $\sid{id} \in \turnouts$, $s\in \internalState$.][h][gr:regulator_program]
	\firstcasesubtil{\ttfamily stmt}{\evGuard{\tid{id}}{n} : \nonterm{A}^\star}{}
	\firstcase{A}{\incr{\bid{id}}}{}
	\otherform{att(\bid{id}, v)}{}
	\otherform{\turn{\sid{id}}{s}}{}
	\otherform{auth}{}
\end{grammar}

\paragraph{Model.}
Overall, states of our model are tuples $\tuple{\trainSeq, \regulator, \switches, \signals}$ where $\trainSeq$ is a set of train tuples, $\regulator$ a regulator tuple, $\switches$ a network configuration (indicating trackworks states) and $\signals$ a signals state function. We suppose all train tuples in $\trainSeq$ have a different identifier. Notice that $\sucblock$ is not part of the state. Instead, the dynamics of the model (see below) is parameterised by $\sucblock$.

@@

In the previous section, we introduced our model. In this section, we present the dynamics of our model, that is the rules that govern how our model can switch from one state to the other. The dynamics is described by means of a reduction relation, characterising the successive states of our model. We define this reduction relation using inference rules, described hereafter.

@@

Train events are not dealt with instantly by the regulator (e.g. two train events can happen before the regulator begins to react to the first one). To characterise this behaviour, train events are stored in a buffer until the regulator uses them. Train events are tuples \tuple{\tid{id}, \bid{id}}, indicating that the train with id $\tid{id}$ enters block \bid{id}. A buffer \bufferFmt{B} is a sequence of such events.

@@

\todo{Rajouté par Guillaume}
A \emph{configuration} is a tuple of the shape $\gamma =  \redTuple{s}{\bufferFmt{B}}{\stateTuple{\trainSeq{}}{\regulator}{\switches}{\signals}}$ with $s$ a guard, $B$ a buffer, and the 4-tuple $\stateTuple{\trainSeq{}}{\regulator}{\switches}{\signals}$ denotes respectively the train sequence, the regulator, the switches and the signals.

\todo{relation: (guard, buffer, état) -> (guard, buffer, etat)}

@@

\begin{mathpar}
	\inferrule*[left=\ruleDef{Start}]{
		\dirFmt{d} \neq \dirFmt{d^\prime} \and P = \trainConcat{\su{d^\prime}{\posFmt{N}}}{P^\prime}
	}{
		\hspace{3mm} \redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{P}\}}{\regulator}{\switches}{\signals}} \\
		\reduces
		\redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d^\prime}{P}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}

Conversely, if the train program is empty (i.e. upon termination), the train stops using rule \ruleRef{Stop}.
\begin{mathpar}
	\inferrule*[left=\ruleDef{Stop}]{
		\dirFmt{d} \neq \dirStop
	}{
	\hspace{4mm}  \redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{\emptyTrainProg}\}}{\regulator}{\switches}{\signals}} \\
	\reduces
	\redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{\dirStop}{\emptyTrainProg}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}

@@

\begin{mathpar}
	\inferrule*[left=\ruleDef{UntilNext}] {
		\suc{\bid{id}}{d}{\switches} = \bid{id^\prime} \neq \nosuc
		\and
		\signalF{\bid{id}}{d} = \siggreen
	}{
		\redTuple{\guardT}{\bufferFmt{B}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id}}{d}{\trainConcat{\su{d}{[\bid{id^\prime}]}}{P}}\}}{\regulator}{\switches}{\signals}} \\
		\reduces		
		\redTuple{\guardS}{\push{\bufferFmt{B}}{\tuple{\tid{id}, \bid{id^\prime}}}}{\stateTuple{\trainSeq{}\cup\{\trainTuple{\tid{id}}{\bid{id^\prime}}{d}{P}\}}{\regulator}{\switches}{\signals}}
	}
\end{mathpar}

@@

\paragraph{Guillaume rajoute la normalisation}

Rule application induce a relation between configuration:  $\gamma \reduces \gamma'$ if $\gamma'$  if there is a rule for which $\gamma$ is the initial state and $\gamma'$ the conclusion. The reflexive-transitive closure of $\reduces$ is denoted $\reduces^*$, that is $\gamma \reduces^* \gamma'$ iff $\gamma = \gamma_0 \reduces \cdots \gamma_n = \gamma'$. The value $n = 0$ corresponds to $\gamma = \gamma'$.

A normal form of a configuration $\gamma$ is a configuration $\gamma'$ such that $\gamma \reduces^* \gamma'$ and there is no $\gamma''$ with $\gamma' \reduces \gamma''$. In other words, starting from $\gamma$, the computation may end on $\gamma'$. We note $\gamma \reduces^! \gamma'$ this relationship.

Given a configuration of the shape  $\gamma = \redTuple{x}{B}{\stateTuple{\trainSeq{}}{\regulator}{\switches}{\signals}}$ and a train $\tid{id}$, 
\begin{itemize}
\item   we say that it has no jobs if  $\trainSeq{}[\tid{id}].P = \emptyTrainProg$;
\item its source is $\mathtt{src}(\gamma, \tid{id}) = \trainSeq{}[\tid{id}].pid$;
\item its destination is $\mathtt{dst}(\gamma, \tid{id}) = pid \in \blocks $ if either
\begin{itemize}
\item  it has no jobs and  $\trainSeq{}[\tid{id}].pid = pid$  or
\item  the last order in  $\trainSeq{}[\tid{id}].P$ is $\su{d}{pid}$ for some direction $d$.
\end{itemize}
\end{itemize}

A configuration $\gamma = \redTuple{\guardT}{\emptyset}{\stateTuple{\trainSeq{}}{\regulator}{\switches}{\signals}} $  is \emph{reliable} if: 
\begin{itemize}
\item  no two trains meet on the same block or the same switch, \todo{En dire plus}
\item $\gamma$ has  a unique normal form $\gamma'$ and 
\item any train in $\gamma'$ has no jobs and 
\item for any train $\tid{id}$, $\mathtt{dst}(\gamma, \tid{id}) =  \mathtt{dst}(\gamma', \tid{id})$. 
\end{itemize}

@@

Variables will be updated with time flying, each transition is described with prime variables. Those latter represent the value of the variable after the transition. For instance, \mintinline{C}{gamma' = [gamma EXCEPT ![T.id].d = "*"]} updates the train tuple: the train \mintinline{C}{T.id}'s direction is set to \mintinline{C}{"*"} where  \mintinline{C}{T} is a variable  defined somewhere else.

To each inference rule in Section~\ref{sec:formal-model}, we associate a TLA+ rule taking the shape (here the Stop rule): 
\begin{minted}{C}
Stop (T) ==
        /\ Len(T.prog) = 0
        /\ meta.G.state = "none"
        /\ T.d /= "*"
        /\ gamma' = [gamma EXCEPT ![T.id].d = "*"]
        /\ meta' = [meta EXCEPT !.msg[1] = Append(meta.msg[1],<<T.id,T.pid>>)]
        /\ UNCHANGED << reg, sigma, F >>
 \end{minted}
The first three lines correspond to the conditions that will fire the rule, the last ones correspond to the configuration update. The next definition represent the application of one rule:
\begin{minted}{C}
Next == 
   \/ UpdateTL
   \/ StartEvent
   \/ Turn
   ...
   \/  \E i \in 1..Len(gamma) :
            \/ Stop(gamma[i])
            \/ Until(gamma[i])
            ...
\end{minted}

The system is then represented by a formula:
\begin{verbatim}
    Spec == Init /\  [][Next]_variables
\end{verbatim}%Revoir la syntaxe au cas où / remplacer par le jolie tla2latex
where variables is the tuple of all variables. The intended meaning is that the initial configuration is specified in \mintinline{C}{Init} and steps follow the \mintinline{C}{Next} formula.  Actually, presently, in our formal semantics, nothing forces a rule to be applied. As a consequence, an ongoing train may never reach its next block. Thanks to TLA+, we can ensure that a rule will eventually be applied (if it can be). Those are known as weak fairness conditions. Again, we specify it as follows. \todo{WF condition?}

@@

The main principle behind our routing process is composition that we discuss now. A route for a train is a sequence $\vec{x} = x_1, \ldots, x_k$ of (contiguous) blocks and switches. Two routes $\vec{x}$ and $\vec{y}$ have a critical resource if  $x_i = y_j$ for some $i,j$. A \emph{global routing} is a route for each train. To compute global routing, we decompose them in  safe steps. A safe step is a global routing such that the routes share no critical resources, neither a block, neither a switch. It is clear that any global routing can be decomposed into such safe steps.

This strategy has a main advantage, it is quite flexible to new events. Suppose that at some point a train is blocked, we may recompute safe steps for the other trains, possibly releasing a deadlock. But, that being said, we are now turned to the transformation of safe steps into train orders and regulator rules.

\section{Expériences}
\label{sec:experiments}
